\documentclass[11pt,a4paper]{article}
\usepackage[top=1.00in, bottom=1.0in, left=1.1in, right=1.1in]{geometry}
\renewcommand{\baselinestretch}{1.2}
\usepackage{graphicx}
\usepackage{natbib}
\usepackage{hyperref}

\newenvironment{smitemize}{
\begin{itemize}
  \setlength{\itemsep}{1pt}
  \setlength{\parskip}{0pt}
  \setlength{\parsep}{0pt}}
{\end{itemize}
}

\def\labelitemi{--}

\begin{document}
\bibliographystyle{/Users/Lizzie/Documents/EndnoteRelated/Bibtex/styles/amnat}
\renewcommand{\refname}{\CHead{}}
% \title{Forecasting Vitis}:

\title{Coordinated phenological responses to environmental cues across a community of temperate forest plants}
\date{13 May 2017}
\author{Flynn \& Wolkovich}
\maketitle

{\bf Extras} \\
Taken from Budburst.Rnw on 15 May 2017:

Temperate woody plants should aim to maximize the carbon gain that comes from starting growth early in the season and gaining first access to critical resources such as light, soil nitrogen and water, while at once minimizing the risk of tissue loss to frost in early spring \cite{Korner:2010,Gu:2008aa,Hufkens:2012aa}. Across a community of co-occuring species we may expect a diversity of strategies to address these pressures---allowing each species to persist in the community. 

It has been found for several cases that short-lived, early successional species typically exhibit such opportunistic strategies, and late-successional species are more typically chilling- and photoperiod-controlled in breaking of dormancy \cite{Korner:2010,Caffarra:2011ab,Basler:2012aa}.

 Latitude of origin (Site) overall had little direct effect on bud burst or leaf-out, but populations from the northern site tended to exhibit slower bud burst and leaf-out, with a more rapid bud burst and leaf out in response to the chilling treatments (indicated by negative coefficients for site x chilling treatments).

Cite Aitken paper on local adaptation....
Need to start to make case here that our study is unique and important!

While nearly all species showed advances in spring phenology in response to the experimental chilling treatment, as indicated by fewer days to phenological events for the 4\degree C and 1.5\degree C treatments, the majority of species (e.g. \emph{Populus grandidentata}) showed delays in both bud burst and leaf out at the more severe chilling treatment. Of the species exposed to the additional chilling, only \emph{Fagus grandifolia} was consistently advanced by the more severe chilling.

It is important to note that the results from the chilling part of this experiment are derived from 11, not 28 species, but the strength of this effect is notable. Strong chilling requirements were detected both for bud burst and leaf out responses, and the most substantial advance in spring phenology came from the more mild chilling treatment, at 4\degree C, with reduced effectiveness of chilling at 1.5\degree C. 

These three factors did show some degree of substituability, meaning for example that a lack of chilling could be made up for by an increase in temperature. These are indicated by the positive two-way interactions; chilling and forcing temperature are more substitutable than chilling and photoperiod, for both bud burst and leaf out.

{\bf Outline}
\begin{enumerate}

\item Introduction
\begin{enumerate}
\item Phenology is super important, to ecosystems and trophic levels and stuff ... and also to coexistence within one trophic level---the temporal niche
\item Climate change has increased research (and interest in phenology) -- lots of change seen, but as data across species has piled up, somewhat conflicting and contrasting results (Fu, chilling etc.). Touch on invasions here? 
\item Accurate predictions clearly require a fuller understanding of the interacting environmental cues that drive phenology within a community. We list out the cues (and say temporal niche again).
\item We somewhere say that we expected a trade-off between cues!
\item Here we study these cues in some forests.
\end{enumerate}

\item Results 
\begin{enumerate}
\item Strong main effects of forcing, chilling and photoperiod. Strong interactions of forcing x chilling; some other important interactions
\item Species varied in their cues, but all species has all 3 cues. Shrubs generally had weaker cues than trees, though this was not always consistent.
\item Responses varied quantitatively depending on whether BB or LO was considered.
\end{enumerate}

\item Discussion
\begin{enumerate}
\item Species and community responses
\begin{enumerate}
\item Spring phenology and community assembly: Our results suggest species have paced budburst and leafout due to a mix of thee major environmental cues. In contrast to any trade-off between cues, we found coordinated responses across the community. This means ...
\item All species had all cues. This jives with many models of phenology (cite Chuine etc.) but diverges from some stated assertions and from a number of chamber studies. Why? Because the cues all interact so if you clip branches once chilling is met or perhaps do the Weinberger thing, you might not see chilling and/or photoperiod cues.
\end{enumerate}
\item Comparing cues
\begin{enumerate}
\item Big effects of forcing and chilling, and photoperiod
\item Chillng at two levels gave similar responses
\item Forcing and chilling offset (cite Fu, Cook etc.) -- implications for climate change 
\item Site had smaller effects, but chilling x site was substantial (raises concerns for single site chilling studies?)
\end{enumerate}
\item Comparing phases: BB vs. LO
\item Conclusions: Tie it all back to climate change
\begin{enumerate}
\item Tie to exotic/non-native work? Tie to pioneer/climax work? 
\item This means predicting any one species' response to climate change may not be simple, let alone predicting community-level responses. 
\end{enumerate}
\end{enumerate}

\end{enumerate}



{\bf Methods:}\\
Total observations, total clippings, total individual trees .... \\
We focused on woody plants, because they can be manipulated in chambers and should be in most direct competition with one another (compared to herbs versus shrubs or such) ....\\



% \newpage
\noindent \emph{Some references}\\

\begin{footnotesize}
{\def\section*#1{}
\bibliography{/Users/Lizzie/Documents/EndnoteRelated/Bibtex/LizzieMainMinimal}
}
\end{footnotesize}

\end{document}

